---
title: "Integration among the three COVID-19 datasets"
date: "`r format(Sys.time(), '%Y %B %d')`"
output:
    html_notebook:
        self_contained: true
        df_print: paged
        fig_height: 6
        fig_width: 6
        highlight: zenburn
        theme: cosmo
        number_sections: yes
        toc: yes
        toc_depth: 6
        toc_float:
            collapsed: false
---

```{r setup}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE
)
```

```{r}
library(tidyverse)
library(DropletUtils)
library(Seurat)
library(SeuratWrappers)
library(harmony)
library(future)
library(ComplexHeatmap)
library(RColorBrewer)
library(scater)
library(edgeR)
library(gghighlight)
library(ggrepel)
library(data.table)
library(ggpubr)
library(scales)
library(fgsea)
library(dendsort)

theme_set(theme_classic(14))

show_dt <- function(df) { 
    DT::datatable(
        df,
        class = "cell-border stripe",
        rownames = TRUE,
        escape = FALSE,
        caption = NULL,
        editable = FALSE,
        filter = list(position = "top"),
        extensions = c('Buttons', 'FixedColumns', 'Scroller'),
        options = list(
            dom = 'Blfrtip',
            scrollX = F,
            scrollY = 500,
            buttons = c('copy', 'csv', 'print'),
            fixedColumns = TRUE,
            deferRender = TRUE,
            scroller = TRUE
        )
    )
}
```

# Load data with cDCs identified in the three COVID-19 datasets

```{r}
dataset1 <- readRDS("mdc_dataset1.rds")
dataset2 <- readRDS("mdc_dataset2.rds")
dataset3 <- readRDS("mdc_dataset3.rds")
```

```{r}
dataset1$condition <- ifelse(dataset1$Sample == 'Covid_1', 'Severe', ifelse(dataset1$Sample %in% c('Covid_2','Covid_3'), 'Mild', 'Healthy'))
dataset1$origin <- 'Dataset 1'
dataset1$patient_id <- paste(dataset1$Sample, 'D1', sep='_')
```

```{r}
dataset2$condition <- ifelse(dataset2$patient_id %in% c('cov01','cov04','cov10','cov11'), 'Severe', ifelse(dataset2$patient_id %in% c('cov02','cov03','cov12'), 'Mild', 'Healthy'))
dataset2$origin <- 'Dataset 2'
dataset2$patient_id <- paste(dataset2$patient_id, 'D2', sep='_')
```

```{r}
dataset3 <- subset(dataset3, subset=orig.ident == 'Covid19') # For dataset 3, keep only COVID19 samples
dataset3$condition <- ifelse(dataset3$group_per_sample == 'severe', 'Severe', 'Mild')
dataset3$origin <- 'Dataset 3'
dataset3$patient_id <- paste(dataset3$donor, 'D3', sep='_')
```

# Merge datasets

```{r}
merged <- merge(dataset1, y=c(dataset2,dataset3), add.cell.ids = c("Dataset1","Dataset2","Dataset3"), project = "cDCs_merged") 

merged <- PercentageFeatureSet(merged, features = grep("^MT\\-", rownames(merged), value = T), col.name = "percent.mt")

VlnPlot(merged, assay = "RNA", slot = "counts", features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), group.by = "origin", same.y.lims = F, log = F, pt.size = 0, ncol = 3)
```
```{r}
merged <- subset(merged, subset = nFeature_RNA >= 500 & percent.mt < 10)
```

```{r}
dim(merged)
```
```{r message=FALSE, warning=FALSE}
merged <- SCTransform(merged, assay = "RNA", new.assay.name = "SCT", do.correct.umi = T, vars.to.regress = c("nCount_RNA","percent.mt"), variable.features.n = 100, return.only.var.genes = F, conserve.memory = F, seed.use = 123) 
merged <- RunPCA(merged, assay = "SCT", npcs = 50)
```

```{r fig.height=4, fig.width=6}
ElbowPlot(merged)
```
```{r}
merged <- RunHarmony(merged, assay.use = "SCT", group.by.vars = c("origin","condition"), reduction = "pca", dims.use = 1:10)
merged <- RunUMAP(merged, reduction = "harmony", dims = 1:10, seed.use = 42)
merged <- FindNeighbors(merged, dims = 1:10, reduction = "harmony", force.recalc = TRUE)
```

```{r}
merged <- FindClusters(merged, algorithm = 1, resolution = 0.8, random.seed = 1)
```

```{r fig.height=4, fig.width=8}
p1 <- DimPlot(merged, reduction = "umap", group.by = c("origin"),  label = F) 
p2 <- DimPlot(merged, reduction = "umap", group.by = c("condition"), label = F, cols = c('#E8867D','#E3BE4C','#D85543'))

p1|p2
```
```{r fig.height=4, fig.width=3}
DimPlot(merged, reduction = "umap", group.by = c("seurat_clusters"), 
        cols = c('9' = '#ED68ED','1' = '#00A9FF', '5' = '#8494FF', '10'='#43BED8','11'='#1667EF',
                 '2' = '#7CAE00', '3' = '#0CB702', '0' = '#0DB423','4' = '#55D112', 
                 '6' = '#04B927','7' = '#61D843','8' = '#43D86D', '12' = '#1BAD44'), label = T) + NoLegend()
```
# Identification of DC subsets

```{r fig.height=10, fig.width=10}
FeaturePlot(merged, reduction = "umap", features = c("CLEC9A","XCR1","CD1C","CLEC10A","FCER1A","CD163","CD14","S100A8","HLA-DRB1"), order = T, min.cutoff = "q05", max.cutoff = "q95", ncol = 3)
```
```{r fig.height=5, fig.width=7}
VlnPlot(merged, features = c("HLA-DRB1", "CLEC10A", "CLEC9A", "CD14", "CD163","S100A8"), pt.size = 0, ncol = 3, cols = c('9' = '#ED68ED','1' = '#00A9FF', '5' = '#8494FF', '10'='#43BED8','11'='#1667EF',
                 '2' = '#7CAE00', '3' = '#0CB702', '0' = '#0DB423','4' = '#55D112', '6' = '#04B927',
                 '7' = '#61D843','8' = '#43D86D', '12' = '#1BAD44'))
```
```{r fig.height=4.5, fig.width=5}
DotPlot(merged, features = c("HLA-DRB1", "CLEC10A", "CLEC9A", "CD14", "CD163","S100A8"), 
        cols = c("lightgrey", "darkgreen"), cluster.idents = T) + RotatedAxis()
```
```{r}
merged$dc <- ifelse(
    merged$seurat_clusters %in% c(1,5,10,11), "DC3", ifelse(
        merged$seurat_clusters %in% c(9), "cDC1", "DC2"
))
```

## cDC proportions 

```{r}
x <- merged[[]]
x <- x[, c('patient_id','condition')]
x <- x[!duplicated(x$patient_id), ]

pt <- table(merged$dc, merged$patient_id)
pt <- as.data.frame(pt)
pt <- merge(pt, x, by.x= "Var2", by.y= "patient_id")
pt <- pt %>% 
  group_by(Var2) %>% 
  mutate(pct=prop.table(Freq))
```

```{r fig.height=3.5, fig.width=5}
ggbarplot(pt, x = "Var1", y = "pct", 
          fill = 'condition',
          palette = c('#E8867D','#E3BE4C','#D85543'),
          add = c("mean_se"),
          position = position_dodge(0.8),
          legend='right')
```
```{r fig.height=3, fig.width=3.5}
ggplot(pt, aes(x = condition, y = pct, fill = Var1)) +
  theme_classic(base_size = 12) +
  geom_col(position = "fill", width = 0.8) +
  xlab("Sample") +
  ylab("Proportion") +
  scale_fill_manual(values =c('#ED68ED','#47B26E','#24C4D2')) +
  theme(legend.title = element_blank())
```
# Re-clustering of DC3 

```{r}
dc3 <- subset(merged, subset=dc=='DC3')
```

```{r message=FALSE, warning=FALSE}
dc3 <- SCTransform(dc3, assay = "RNA", new.assay.name = "SCT", do.correct.umi = T, vars.to.regress = c("nCount_RNA","percent.mt"), variable.features.n = 100, return.only.var.genes = F, conserve.memory = F, seed.use = 123) 
dc3 <- RunPCA(dc3, assay = "SCT", npcs = 50) 
```

```{r fig.height=4, fig.width=6}
ElbowPlot(dc3, ndims = 30)
```
```{r}
dc3 <- RunHarmony(dc3, assay.use = "SCT", group.by.vars = c("origin", "condition"), reduction = "pca", dims.use = 1:30)
dc3 <- RunUMAP(dc3, reduction = "harmony", dims = 1:30, seed.use = 42)
dc3 <- FindNeighbors(dc3, dims = 1:30, reduction = "harmony", force.recalc = TRUE)
dc3 <- FindClusters(dc3, algorithm = 1, resolution = 0.6, random.seed = 1)
```

```{r}
dim(dc3)
```
```{r fig.height=4, fig.width=3.5}
DimPlot(dc3, reduction = "umap", group.by = 'seurat_clusters', label = T, 
        cols = c('0' = '#BBA8A8', '2' = '#9C9595', '1' = '#C6CCC7', '3' ='#E1E1E1','4' = '#B1B3B0')) + NoLegend()
```
```{r fig.height=3, fig.width=4}
DotPlot(dc3, features = c("CD14", "CD163"), 
        cols = c("lightgrey", "darkgreen"), cluster.idents = T) + RotatedAxis()
```
```{r}
dc3$annotation <- ifelse(
    dc3$seurat_clusters %in% c(1,3,4), "CD14low CD163low", "CD14+ CD163+")
```

## DC3 proportions

```{r}
x <- dc3[[]]
x <- x[, c('patient_id','condition')]
x <- x[!duplicated(x$patient_id), ]

pt <- table(dc3$annotation, dc3$patient_id)
pt <- as.data.frame(pt)
pt <- merge(pt, x, by.x= "Var2", by.y= "patient_id")
pt <- pt %>% 
  group_by(Var2) %>% 
  mutate(pct=prop.table(Freq))

pt$disease <- ifelse(pt$condition %in% c('Mild','Severe'), 'COVID-19','HD')
```

```{r fig.height=3, fig.width=4}
ggplot(pt, aes(x = factor(disease,levels=c('HD','COVID-19')), y = pct, fill = factor(Var1, levels=c('CD14low CD163low', 'CD14+ CD163+')))) +
  theme_classic(base_size = 12) +
  geom_col(position = "fill", width = 0.8) +
  xlab("Sample") +
  ylab("Proportion") +
  scale_fill_manual(values =c('#C6CCC7','#BBA8A8')) +
  theme(legend.title = element_blank())
```
# Direct comparison COVID19 vs Healthy DC3 vs DC2 

## DEGs

Remove donors with less than 10 cells in DC2 or DC3.

```{r}
sce <- as.SingleCellExperiment(merged, assay = "SCT")
sce <- sce[rowSums(counts(sce) > 2) >= 10, ]
```

```{r}
t(table(sce$dc, sce$patient_id))
t(table(sce$disease, sce$patient_id))
```
```{r}
sce_dc23 <- sce[, (sce$patient_id %in% c('C19-CB-0001_D3','C19-CB-0002_D3','C19-CB-0003_D3','C19-CB-0009_D3','cov01_D2','cov03_D2','cov07_D2','cov08_D2','cov09_D2','cov11_D2','cov12_D2','cov17_D2','cov18_D2','Covid_2_D1','Covid_3_D1','HD_1_D1','HD_2_D1')) & (!sce$dc == "cDC1")]
t(table(sce_dc23$dc, sce_dc23$patient_id))
```
```{r}
sce_dc23$cluster_id <- factor(sce_dc23$dc, levels = c("DC2","DC3"))
sce_dc23$group_id <- factor(sce_dc23$disease, levels = c("covid","healthy"))
sce_dc23$sample_id <- factor(sce_dc23$patient_id)

levels(sce_dc23$sample_id)[levels(sce_dc23$sample_id)=="C19-CB-0001_D3"] <- "C19_CB_0001_D3"
levels(sce_dc23$sample_id)[levels(sce_dc23$sample_id)=="C19-CB-0002_D3"] <- 'C19_CB_0002_D3'
levels(sce_dc23$sample_id)[levels(sce_dc23$sample_id)=="C19-CB-0003_D3"] <- 'C19_CB_0003_D3'
levels(sce_dc23$sample_id)[levels(sce_dc23$sample_id)=="C19-CB-0009_D3"] <- 'C19_CB_0009_D3'

sce_dc23$cid_gid <- factor(paste(sce_dc23$cluster_id, sce_dc23$group_id, sep = "_"))

dge_dc23 <- aggregateAcrossCells(sce_dc23, id = colData(sce_dc23)[, c("sample_id","cluster_id")])
dge_dc23 <- calcNormFactors(dge_dc23)

mtx_dc23 <- t(scale(t(edgeR::cpm(dge_dc23, log = T))))
colnames(mtx_dc23) <- paste(dge_dc23$samples$cid_gid, dge_dc23$samples$sample_id, sep = "_")

dsn_dc23 <- model.matrix(~ sample_id, dge_dc23$samples)
dsn_dc23 <- cbind(
    dsn_dc23,
    DC3_healthy = dge_dc23$samples$cluster_id == "DC3" & dge_dc23$samples$group_id == "healthy",
    DC3_covid = dge_dc23$samples$cluster_id == "DC3" & dge_dc23$samples$group_id == "covid"
)

dge_dc23 <- estimateDisp(dge_dc23, dsn_dc23)
fit_dc23 <- glmQLFit(dge_dc23, dsn_dc23, robust = T)

cts_dc23 <- makeContrasts(
    healthy.DC3_vs_DC2 = DC3_healthy,
    covid.DC3_vs_DC2 = DC3_covid,
    covid_vs_healthy.DC3_vs_DC2 = DC3_covid - DC3_healthy,
    levels = dsn_dc23
)

res_dc23 <- glmQLFTest(fit_dc23, contrast = cts_dc23[, "covid_vs_healthy.DC3_vs_DC2"]) %>%
    topTags(n = Inf) %>% 
    `[[`("table") %>%
    `[`(c("logFC", "PValue", "FDR")) %>%
    rownames_to_column("gene") %>%
    as_tibble()
```

```{r}
res_dc23 %>%
    show_dt()
```

### Volcano plot 

```{r}
res_dc23$threshold <- "NoVariance"
res_dc23$threshold[res_dc23$PValue < 0.05] <- c("-1 < FC < 1")
res_dc23$threshold[res_dc23$PValue < 0.05 & res_dc23$logFC > 1] <- c("FC > 1")
res_dc23$threshold[res_dc23$PValue < 0.05 & res_dc23$logFC < -1] <- c("FC < -1")
```

```{r}
res_dc23$threshold<-factor(res_dc23$threshold)
res_dc23$threshold<-factor(res_dc23$threshold, levels(res_dc23$threshold)[c(4,2,1,3)])

res_dc23$labels <- with(res_dc23, ifelse(gene %in% c('CLU','AXL','C1QA','CD14')|logFC < -1.5 & PValue < 0.05, gene, ""))
```

```{r fig.height=4, fig.width=5.5}
ggplot(data=res_dc23, aes(x=logFC, y=-log10(PValue), colour=threshold)) +
  geom_point(alpha=0.8, size=1.75) +
  xlim(c(-6.2,6.2))+
  ylim(c(0, 5)) +
  xlab("log2 Fold Change") + ylab("-log10(p-value)") +
  scale_color_manual(values=c("darkgrey","blue","lightgrey", "red")) +
  geom_vline(xintercept = -1,colour="blue", linetype = "longdash") +
  geom_vline(xintercept = 1,colour="red", linetype = "longdash") +
  geom_hline(yintercept = -log10(0.05),colour="blue", linetype = "longdash") +
  geom_text_repel(aes(label=labels), size = 4) +
  theme(panel.grid.major = element_line(colour = "white"),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA, colour = "black", size = 1),
        legend.position = "none", legend.key = element_blank(), axis.text = element_text(size=13))
```
# Severe vs Mild COVID-19 patients

Aggregate DCs by type and donor, then run differential test on aggregated counts. This ensures that the unit of independent replication is the sample (donor), not the cell.

```{r}
t(table(sce$dc, sce$patient_id))
t(table(sce$condition, sce$patient_id))
```
## DC2

22 donors have more than 10 DC2s (7 HDs, 9 mild, 6 severe). 

### DEGs: Severe vs Mild

```{r}
sce_dc2 <- sce[, (sce$patient_id %in% c('C19-CB-0001_D3','C19-CB-0002_D3','C19-CB-0003_D3','C19-CB-0009_D3','C19-CB-0021_D3','C19-CB-0198_D3','C19-CB-0204_D3','cov01_D2','cov02_D2','cov03_D2','cov04_D2','cov07_D2','cov08_D2','cov09_D2','cov11_D2','cov12_D2','cov17_D2','cov18_D2','Covid_2_D1','Covid_3_D1','HD_1_D1','HD_2_D1')) & (sce$dc == "DC2")]
t(table(sce_dc2$condition, sce_dc2$patient_id))
```
```{r}
sce_dc2$group_id <- factor(sce_dc2$condition, levels = c("Healthy", "Mild", "Severe"))
sce_dc2$sample_id <- factor(sce_dc2$patient_id)
sce_dc2$dataset <- factor(sce_dc2$origin)

levels(sce_dc2$dataset)[levels(sce_dc2$dataset)=="Dataset 1"] <- "Dataset1"
levels(sce_dc2$dataset)[levels(sce_dc2$dataset)=="Dataset 2"] <- "Dataset2"
levels(sce_dc2$dataset)[levels(sce_dc2$dataset)=="Dataset 3"] <- "Dataset3"

dge_dc2 <- aggregateAcrossCells(sce_dc2, id = colData(sce_dc2)[, c("sample_id")])
dge_dc2 <- calcNormFactors(dge_dc2)

mtx_dc2 <- t(scale(t(edgeR::cpm(dge_dc2, log = T))))

dsn_dc2 <- model.matrix(~ dataset + group_id, dge_dc2$samples)
dge_dc2 <- estimateDisp(dge_dc2, dsn_dc2)
fit_dc2 <- glmQLFit(dge_dc2, dsn_dc2, robust = T)

cts_dc2 <- makeContrasts(
    DC2_severe_vs_mild = group_idSevere - group_idMild,
    levels = dsn_dc2
)

res_dc2 <- glmQLFTest(fit_dc2, contrast = cts_dc2[, "DC2_severe_vs_mild"]) %>%
    topTags(n = Inf) %>% 
    `[[`("table") %>%
    `[`(c("logFC", "PValue", "FDR")) %>%
    rownames_to_column("gene") %>%
    as_tibble()
```

```{r}
res_dc2 %>%
    show_dt()
```

## DC3

19 donors have more than 10 DC2s (7 HDs, 7 mild, 5 severe). 

### DEGs: Severe vs Mild

```{r}
sce_dc3 <- sce[, (sce$patient_id %in% c('C19-CB-0001_D3','C19-CB-0002_D3','C19-CB-0003_D3',
'C19-CB-0008_D3', 'C19-CB-0009_D3','cov01_D2','cov03_D2','cov07_D2','cov08_D2','cov09_D2','cov11_D2','cov12_D2','cov17_D2','cov18_D2','Covid_1_D1','Covid_2_D1', 'Covid_3_D1','HD_1_D1','HD_2_D1')) & (sce$dc == "DC3")]
t(table(sce_dc3$condition, sce_dc3$patient_id))
```
```{r}
sce_dc3$group_id <- factor(sce_dc3$condition, levels = c("Healthy", "Mild", "Severe"))
sce_dc3$sample_id <- factor(sce_dc3$patient_id)
sce_dc3$dataset <- factor(sce_dc3$origin)

levels(sce_dc3$dataset)[levels(sce_dc3$dataset)=="Dataset 1"] <- "Dataset1"
levels(sce_dc3$dataset)[levels(sce_dc3$dataset)=="Dataset 2"] <- "Dataset2"
levels(sce_dc3$dataset)[levels(sce_dc3$dataset)=="Dataset 3"] <- "Dataset3"

dge_dc3 <- aggregateAcrossCells(sce_dc3, id = colData(sce_dc3)[, c("sample_id")])
dge_dc3 <- calcNormFactors(dge_dc3)

mtx_dc3 <- t(scale(t(edgeR::cpm(dge_dc3, log = T))))

dsn_dc3 <- model.matrix(~ dataset + group_id, dge_dc3$samples)
dge_dc3 <- estimateDisp(dge_dc3, dsn_dc3)
fit_dc3 <- glmQLFit(dge_dc3, dsn_dc3, robust = T)

cts_dc3 <- makeContrasts(
    DC3_severe_vs_mild = group_idSevere - group_idMild,
    levels = dsn_dc3
)

res_dc3 <- glmQLFTest(fit_dc3, contrast = cts_dc3[, "DC3_severe_vs_mild"]) %>%
    topTags(n = Inf) %>% 
    `[[`("table") %>%
    `[`(c("logFC", "PValue", "FDR")) %>%
    rownames_to_column("gene") %>%
    as_tibble()
```

```{r}
res_dc3 %>%
    show_dt()
```

# Volcano plot Severe vs Mild in DC2 and DC3 

## DC2 

```{r}
res_dc2$threshold <- "NoVariance"
res_dc2$threshold[res_dc2$PValue < 0.05] <- c("-1 < FC < 1")
res_dc2$threshold[res_dc2$PValue < 0.05 & res_dc2$logFC > 1] <- c("FC > 1")
res_dc2$threshold[res_dc2$PValue < 0.05 & res_dc2$logFC < -1] <- c("FC < -1")

res_dc2$threshold<-factor(res_dc2$threshold)
res_dc2$threshold<-factor(res_dc2$threshold, levels(res_dc2$threshold)[c(4,2,1,3)])
```

```{r}
res_dc2$labels <- with(res_dc2, ifelse((abs(logFC) > 2 & PValue < 0.05)|res_dc2$gene %in% c("C1QB","THBD","MT1E"), gene, ""))
```

```{r fig.height=5, fig.width=7}
ggplot(data=res_dc2, aes(x=logFC, y=-log10(PValue), colour=threshold)) +
  geom_point(alpha=0.8, size=1.75) +
  xlim(c(-6,6))+
  ylim(c(0, 6.5)) +
  xlab("log2 Fold Change") + ylab("-log10(p-value)") +
  scale_color_manual(values=c("darkgrey","blue","lightgrey", "red")) +
  geom_vline(xintercept = -1,colour="blue", linetype = "longdash") +
  geom_vline(xintercept = 1,colour="red", linetype = "longdash") +
  geom_hline(yintercept = -log10(0.05),colour="blue", linetype = "longdash") +
  geom_text_repel(aes(label=labels), size=4.5) +
  theme(panel.grid.major = element_line(colour = "white"),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA, colour = "black", size = 1),
        legend.position = "none", legend.key = element_blank())
```
## DC3

```{r}
res_dc3$threshold <- "NoVariance"
res_dc3$threshold[res_dc3$PValue < 0.05] <- c("-1 < FC < 1")
res_dc3$threshold[res_dc3$PValue < 0.05 & res_dc3$logFC > 1] <- c("FC > 1")
res_dc3$threshold[res_dc3$PValue < 0.05 & res_dc3$logFC < -1] <- c("FC < -1")

res_dc3$threshold<-factor(res_dc3$threshold)
res_dc3$threshold<-factor(res_dc3$threshold, levels(res_dc3$threshold)[c(4,2,1,3)])
```

```{r}
res_dc3$labels <- with(res_dc3, ifelse((abs(logFC) > 2 & PValue < 0.05)|res_dc3$gene %in% c("S100A8","C5AR1","ALOX5AP","THBS1","CD14","S100A9","S100A12","ADAM9","CLU"), gene, ""))
res_dc3$labels <- with(res_dc3, ifelse(res_dc3$labels %in% c("ENSG00000276216","ENSG00000259001"), " ", res_dc3$labels))
```

```{r fig.height=5, fig.width=7}
ggplot(data=res_dc3, aes(x=logFC, y=-log10(PValue), colour=threshold)) +
  geom_point(alpha=0.8, size=1.75) +
  xlim(c(-6,6))+
  ylim(c(0, 6.5)) +
  xlab("log2 Fold Change") + ylab("-log10(p-value)") +
  scale_color_manual(values=c("darkgrey","blue","lightgrey", "red")) +
  geom_vline(xintercept = -1,colour="blue", linetype = "longdash") +
  geom_vline(xintercept = 1,colour="red", linetype = "longdash") +
  geom_hline(yintercept = -log10(0.05),colour="blue", linetype = "longdash") +
  geom_text_repel(aes(label=labels), size=4.5) +
  theme(panel.grid.major = element_line(colour = "white"),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA, colour = "black", size = 1),
        legend.position = "none", legend.key = element_blank())

```
# Boxplots of selected genes

## DC2

```{r}
genes <- c('HLA-DRA','HLA-DQA2','HLA-DPB1','HLA-DPA1','CD86','IL1B','CCL3','CCL4')

mtx_dc2.cpm <- edgeR::cpm(dge_dc2, log = T)
mtx_dc2.cpm <- as.data.frame(mtx_dc2.cpm)
mtx_dc2.cpm <- mtx_dc2.cpm[genes, ]

colnames(mtx_dc2.cpm) <- paste(dge_dc2$samples$condition, dge_dc2$samples$patient_id, sep = "_")
mtx_dc2.cpm <- t(mtx_dc2.cpm)
mtx_dc2.cpm <- as.data.frame(mtx_dc2.cpm)
```

```{r}
mtx_dc2.cpm$condition <- ifelse(grepl('^Healthy', rownames(mtx_dc2.cpm)),'HD', ifelse(
    grepl('^Mild', rownames(mtx_dc2.cpm)),'Mild','Severe'))
mtx_dc2.cpm <- with(mtx_dc2.cpm, mtx_dc2.cpm[order(condition) , ])
```

```{r}
my_comparisons <- list(c("HD", "Mild"), c("Mild", "Severe"), c("HD", "Severe"))

my_plot_list <- list()
for(i in 1:length(genes)){
  p <- ggboxplot(mtx_dc2.cpm, x = "condition", y = genes[i], 
                 color = "black", fill=c('pink', 'lightblue','blue'), legend = "none", 
                 ylab = "Expression level", xlab = FALSE, title=genes[i], width = 0.8) +
      stat_compare_means(aes(label = ..p.format..), comparisons = my_comparisons) + 
      theme(plot.title = element_text(hjust = 0.5, size=14), axis.text.x = element_blank(), axis.title = element_blank())
                
  my_plot_list[[i]] <- p
}
```

```{r fig.height=3.8, fig.width=18}
my_plot_list[[1]]|my_plot_list[[2]]|my_plot_list[[3]]|my_plot_list[[4]]|my_plot_list[[5]]|my_plot_list[[6]]|my_plot_list[[7]]|my_plot_list[[8]]
```

## DC3

```{r}
mtx_dc3.cpm <- edgeR::cpm(dge_dc3, log = T)
mtx_dc3.cpm <- as.data.frame(mtx_dc3.cpm)
mtx_dc3.cpm <- mtx_dc3.cpm[genes, ]

colnames(mtx_dc3.cpm) <- paste(dge_dc3$samples$condition, dge_dc3$samples$patient_id, sep = "_")
mtx_dc3.cpm <- t(mtx_dc3.cpm)
mtx_dc3.cpm <- as.data.frame(mtx_dc3.cpm)
```

```{r}
mtx_dc3.cpm$condition <- ifelse(grepl('^Healthy', rownames(mtx_dc3.cpm)),'HD', ifelse(
    grepl('^Mild', rownames(mtx_dc3.cpm)),'Mild','Severe'))
mtx_dc3.cpm <- with(mtx_dc3.cpm, mtx_dc3.cpm[order(condition) , ])
```

```{r}
my_plot_list <- list()
for(i in 1:length(genes)){
  p <- ggboxplot(mtx_dc3.cpm, x = "condition", y = genes[i], 
                 color = "black", fill=c('pink', 'lightblue','blue'), legend = "none", 
                 ylab = "Expression level", xlab = FALSE,title=genes[i], width = 0.8) +
      stat_compare_means(aes(label = ..p.format..), comparisons = my_comparisons) + 
      theme(plot.title = element_text(hjust = 0.5, size=14), axis.text.x = element_blank(), axis.title = element_blank())
                
  my_plot_list[[i]] <- p
}
```

```{r fig.height=3.8, fig.width=18}
my_plot_list[[1]]|my_plot_list[[2]]|my_plot_list[[3]]|my_plot_list[[4]]|my_plot_list[[5]]|my_plot_list[[6]]|my_plot_list[[7]]|my_plot_list[[8]]
```
# GSEA with Hallmark collection

```{r}
gmt.hallmark <- gmtPathways("h.all.v7.2.symbols.gmt")
```

## DC2

```{r}
rnk_dc2 <- setNames(-log10(res_dc2$PValue) * sign(res_dc2$logFC), res_dc2$gene)
gsa_dc2.hallmark <- fgseaMultilevel(gmt.hallmark, rnk_dc2, minSize = 10, maxSize = 500, sampleSize = 100)
```

```{r}
gsa_dc2.hallmark %>%
    as_tibble() %>%
    arrange(pval) %>%
    dplyr::select(pathway,NES, pval,padj,leadingEdge) %>%
    show_dt()
```

## DC3

```{r}
rnk_dc3 <- setNames(-log10(res_dc3$PValue) * sign(res_dc3$logFC), res_dc3$gene)
gsa_dc3.hallmark <- fgseaMultilevel(gmt.hallmark, rnk_dc3, minSize = 10, maxSize = 500, sampleSize = 100)
```

```{r}
gsa_dc3.hallmark %>%
    as_tibble() %>%
    arrange(pval) %>%
    dplyr::select(pathway,NES, pval,padj,leadingEdge) %>%
    show_dt()
```

# Pathway enrichment visualization

For each cDC subset, top 15 pathways (ordered by pvalue) are selected, merged and shown.

```{r}
gsa_dc2.hallmark <- arrange(gsa_dc2.hallmark, pval)
gsa_dc2.hallmark_top <- gsa_dc2.hallmark[1:15, ]

gsa_dc3.hallmark <- arrange(gsa_dc3.hallmark, pval)
gsa_dc3.hallmark_top <- gsa_dc3.hallmark[1:15, ] 

pathways.hallmark <- rbind(gsa_dc2.hallmark_top, gsa_dc3.hallmark_top)
pathways.hallmark2 <- pathways.hallmark[!duplicated(pathways.hallmark$pathway),]
pathways.hallmark2 <- pathways.hallmark2[, -c(2:8)]

result_dc2 <- merge(pathways.hallmark2, gsa_dc2.hallmark)
result_dc2$DC_subsets <- 'DC2'
result_dc2$count <- -log10(result_dc2$pval) 

result_dc3 <- merge(pathways.hallmark2, gsa_dc3.hallmark)
result_dc3$DC_subsets <- 'DC3'
result_dc3$count <- -log10(result_dc3$pval) 

hallmark <- rbind(result_dc2, result_dc3)
```

```{r}
hallmark$pathway <- gsub('HALLMARK_', '', hallmark$pathway)
hallmark$pathway <- gsub('_', ' ', hallmark$pathway)
```

```{r}
order <- setDT(hallmark)[, lapply(.SD, sum), by = pathway, .SDcols = 'NES']
names(order)[names(order) == "NES"] <- "rowwise_sum_NES"
hallmark <- merge(x = hallmark, y = order, by='pathway', all.x = TRUE)
```

```{r fig.height=3.5, fig.width=4.5}
p <- ggplot(hallmark, aes(x = DC_subsets, y = fct_reorder(pathway,rowwise_sum_NES))) +
    geom_point(aes(size = count, colour = NES)) +
    ylab(NULL) + xlab(NULL) + 
    scale_color_gradientn(colors = c("darkblue", "lightblue", "white", "pink", "darkred"),
                    values = scales::rescale(c(-2.2, -1.3, -1, 1, 1.3, 2.2)),
                    limits = c(-2.2, 2.2)) +
    labs(size = "-Log10(PValue)") +
    theme(axis.text.x = element_text(angle = 00, hjust = 0.5, size=8), 
          axis.text.y = element_text(size=7),
          legend.title = element_text(size = 9),
          legend.key.size = unit(0.2, "cm"),
          legend.key.width = unit(0.2,"cm"), 
          legend.text=element_text(size=5))
p
```
```{r}
ggsave("HALLMARK_severe_vs_mild.png", p, width = 4.2, height = 3, units = c("in"), dpi=600)
```

# Heatmaps with GSEA leading edge genes of the allograft rejection pathway

## DC2 severe vs mild

```{r fig.height=5.5, fig.width=3.5}
colnames(mtx_dc2) <- paste(dge_dc2$samples$group_id, dge_dc2$samples$sample_id, sep = "_")

mtx_dc2[gsa_dc2.hallmark[pathway == "HALLMARK_ALLOGRAFT_REJECTION", leadingEdge][[1]], c(grep("Healthy",colnames(mtx_dc2)), grep("Mild",colnames(mtx_dc2)), grep("Severe", colnames(mtx_dc2)) )] %>%
    Heatmap(
        cluster_rows = T, 
        cluster_columns = F,
        clustering_method_rows = "ward.D2", 
        row_names_gp = gpar(fontsize = 6.5),
        col = circlize::colorRamp2(seq(-2, 2, length.out = 11), rev(brewer.pal(11, "RdYlBu")))
    )
```
## DC3 severe vs mild

```{r fig.height=5.5, fig.width=3.5}
colnames(mtx_dc3) <- paste(dge_dc3$samples$group_id, dge_dc3$samples$sample_id, sep = "_")

mtx_dc3[gsa_dc3.hallmark[pathway == "HALLMARK_ALLOGRAFT_REJECTION", leadingEdge][[1]], c(grep("Healthy",colnames(mtx_dc3)),grep("Mild",colnames(mtx_dc3)), grep("Severe", colnames(mtx_dc3)) )] %>%
    Heatmap(
        cluster_rows = T, 
        cluster_columns = F,
        clustering_method_rows = "ward.D2", 
        row_names_gp = gpar(fontsize = 6.5),
        col = circlize::colorRamp2(seq(-2, 2, length.out = 11), rev(brewer.pal(11, "RdYlBu")))
    )
```
